Dirace: Practical Proof Automation of Dirac Notation Equations

### 2025-01-14 labelled.tex
2025/1/14

\begin{gather}
  \textbf{W-AssumE-Reg}\qquad 
  \frac{E[]\vdash \sigma : \Index\qquad r\notin E}{\WF[E; r : \reg(\sigma)[]]}
\end{gather}

\begin{gather*}
  \textbf{Type-Labelled} \qquad
  \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} \\
  \\
  \textbf{Reg-Var} \qquad
  \frac{\WF(E[\Gamma]) \qquad r : \reg(\sigma) \in E}{E[\Gamma] \vdash r : \reg(\sigma)}  \\ 
  \\
  \textbf{Reg-Pair}\qquad
  \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash Q : \reg \tau\qquad R\| Q}{E[\Gamma] \vdash (R,Q) : \reg \sigma \times \tau}
\end{gather*}

\begin{gather*}
    \textbf{Type-L-Base-Ket}\qquad 
    \frac{r : \reg \sigma \in E\qquad E[\Gamma] \vdash i : \Basis(\sigma)}{E[\Gamma] \vdash |i\>_r : \DType(\{r\}; \emptyset)} \\
    \textbf{Type-L-Base-Bra}\qquad 
    \frac{r : \reg \sigma \in E \qquad E[\Gamma] \vdash i : \Basis(\sigma)}{E[\Gamma] \vdash {}_r\<i| : \DType(\emptyset; \{r\})} \\
    \textbf{Type-L-Ket}\qquad 
    \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R; \emptyset)} \\
    \textbf{Type-L-Bra}\qquad 
    \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B_R : \DType(\emptyset; \var R)} \\
    \textbf{Type-L-Opt}\qquad 
    \frac{E[\Gamma] \vdash R_1 : \reg \sigma_1\qquad 
    E[\Gamma] \vdash R_2 : \reg \sigma_2\qquad 
    E[\Gamma] \vdash O : \OType(\sigma_1; \sigma_2)}
    {E[\Gamma] \vdash O_{R_1;R_2} : \DType(\var R_1; \var R_2)} \\
    \textbf{Type-L-Conj}\qquad 
    \frac{E[\Gamma] \vdash \tD : \DType(s_1;s_2)}{E[\Gamma] \vdash \tD^\dagger : \DType(s_2;s_1)} \qquad
    \textbf{Type-L-SCL}\qquad 
    \frac{E[\Gamma] \vdash S : \SType\qquad E[\Gamma] \vdash \tD : \DType(s_1;s_2)}{E[\Gamma] \vdash S.\tD : \DType(s_1;s_2)} \\
    \textbf{Type-L-ADD}\qquad
    \frac{E[\Gamma] \vdash \tD_i : \DType(s_1;s_2)\quad \text{forall } i}{E[\Gamma] \vdash \tD_1+\cdots+\tD_n : \DType(s_1;s_2)} \\
    \textbf{Type-L-TSR}\qquad
    \frac{E[\Gamma] \vdash \tD_1 : \DType(s_1;s_1')\qquad E[\Gamma] \vdash \tD_2 : \DType(s_2;s_2')\qquad s_1\cap s_2 = \emptyset\qquad s_1'\cap s_2' = \emptyset}{E[\Gamma] \vdash \tD_1\otimes\tD_2 : \DType(s_1\cup s_2;s_1'\cup s_2')} \\
    \textbf{Type-L-DOT}\qquad
    \frac{E[\Gamma] \vdash \tD_1 : \DType(s_1;s_1')\qquad E[\Gamma] \vdash \tD_2 : \DType(s_2;s_2')\qquad 
    s_1 \cap s_2 \backslash s_1' = \emptyset \qquad
    s_2' \cap s_1' \backslash s_2 = \emptyset}
    {E[\Gamma] \vdash \tD_1\cdot\tD_2 : \DType(s_1 \cup (s_2\backslash s_1'); s_2' \cup (s_1'\backslash s_2))}
\end{gather*}

@@

\begin{enumerate}
  \item For any register $R$ s.t. $E[\Gamma]\vdash R : \reg \sigma$, suppose $\var R = \{r_1,r_2,\cdots,r_n\}$, 
  we introduce variables $i_{r_k} : \Basis(\sigma_{r_k})$ with $r_k : \reg \sigma_{r_k} \in E$ for $k = 1,\cdots, n$. We reconstruct the basis $|i_R\>$ (which is of type $\KType(\sigma)$) and $\<i_R|$ (which is of type $\BType(\sigma)$) of $R$ by:
  \begin{itemize}
    \item $R = r_k$, $|i_R\> \triangleq |i_{r_k}\>$ and $\<i_R| \triangleq \<i_{r_k}|$;
    \item $R = (R_1,R_2)$: $|i_R\> \triangleq |i_{R_1}\> \otimes |i_{R_2}\>$ and $\<i_R| \triangleq \<i_{R_1}| \otimes \<i_{R_2}|$.
  \end{itemize}
    \item Expansion of all $K_R,B_R,O_{R,R'}$:
    \begin{align*}
      &K_R = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (\<i_R|\cdot K). (|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n}).\\
      &B_R = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (B\cdot |i_R\>). ({}_{r_1}\<i_{r_1}|\otimes\cdots\otimes{}_{r_n}\<i_{r_n}|).\\
      &O_{R,R'} = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
      \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
      &\qquad\qquad\qquad (\<i_R|\cdot O\cdot |i_{R'}\>).\Big((|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n})\otimes({}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|)\Big).\\
    \end{align*}
    \item Rewriting to push big operators and split: \textbf{Beta-Arrow}, \textbf{Beta-Index}, \textbf{Delta}

\begin{ruletable}{The special flattening rule.}
    %
    R-ADJDK
    & $ ({}_r\<i|)^\dagger \ \reduce\ |i\>_r$ \\
    %
    R-ADJDB
    & $ (|i\>_r)^\dagger \ \reduce\ {}_r\<i|$ \\
    & \textcolor{gray}{Dealing with adjoint; now we don't have any $\tD^\dagger$ by combining push rules.} \\
    R-ADJD0
    & $ (\tD_1 \otimes \tD_2)^\dagger \ \reduce\ \tD_1^\dagger \otimes \tD_2^\dagger$ \\
    %
    R-ADJD1
    & $ (\tD_1\cdot \tD_2)^\dagger \ \reduce\ \tD_2^\dagger \cdot \tD_1^\dagger$ \\
    %
    R-SCRD0
    & $ D_1 \otimes \cdots \otimes (a.\tD_n) \otimes \cdots \otimes \tD_m \ \reduce\ a.(\tD_1 \otimes \cdots \otimes \tD_m) $ \\
    %
    R-SCRD1
    & $ (a.\tD_1) \cdot \tD_2 \ \reduce\ a.(\tD_1 \cdot \tD_2) $ \\
    %
    R-SCRD2
    & $ \tD_1 \cdot (a.\tD_2) \ \reduce\ a.(\tD_1 \cdot \tD_2) $ \\
    & \textcolor{gray}{Dealing with scaling} \\
    %
    R-TSRD0
    & $ (\tD_1 + \cdots + \tD_n) \otimes \tD\ \reduce\ \tD_1 \otimes \tD + \cdots + \tD_n \otimes \tD $ \\
    %
    R-DOTD0
    & $ (\tD_1 + \cdots + \tD_n) \cdot \tD\ \reduce\ \tD_1 \cdot \tD + \cdots + \tD_n \cdot \tD $ \\
    %
    R-DOTD1
    & $ \tD \cdot (\tD_1 + \cdots + \tD_n)\ \reduce\ \tD \cdot \tD_1 + \cdots + \tD \cdot \tD_n $ \\
    & \textcolor{gray}{Distributivity of $\cdot$ and $\otimes$. } \\
    %
    R-SUM-PUSHD0
    & $ (\sum_{i \in M} \tD_1) \otimes \tD_2 \ \reduce\ \sum_{i \in M} (\tD_1 \otimes \tD_2) $ \\
    %
    R-SUM-PUSHD2
    & $ (\sum_{i \in M} \tD_1) \cdot \tD_2 \ \reduce\ \sum_{i \in M} (\tD_1 \cdot \tD_2) $ \\
    %
    R-SUM-PUSHD3
    & $ \tD_1 \cdot (\sum_{i \in M} \tD_2) \ \reduce\ \sum_{i \in M} (\tD_1 \cdot \tD_2) $ \\
    %
    R-L-SORT0
    & $ A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B $ \\
    %
    R-L-SORT1
    & ${}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j}$ \\
    %
    R-L-SORT2
    & ${}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)$ \\
    %
    R-L-SORT3
    & $(X_1 \otimes \cdots \otimes {}_r\bra{i} \otimes \cdots \otimes X_n) \cdot \ket{j}_r \ \reduce\ \delta_{i,j}.(X_1 \otimes \cdots \otimes X_n)$ \\
    %
    R-L-SORT1
    & $ (X_1 \otimes \cdots \otimes {}_r\bra{i} \otimes \cdots \otimes X_n) \cdot (Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) $ \\
    & $\reduce\ \delta_{i,j}.(X_1 \otimes \cdots \otimes X_n) \cdot (Y_1 \otimes \cdots \otimes Y_m)$
\end{ruletable}
    After the rewriting of this step, the expression we have is always in the following form:
    $$
    \Big(\sum_{i_1}\cdots\sum_{i_n} a.\tD_f\Big) + \cdots + \Big(\sum_{i_1}\cdots\sum_{i_{n'}} a.\tD_f\Big)
    $$
    where $\tD_f$ is generated by :
    $$\tD_f ::= |i\>_r \mid {}_r\<i| \mid \te\otimes\te \mid \te\cdot\te. $$
    \item Simplify $\tD_f$. Let $\cN$ be the variable name set. For every $\tD_f$, we define three functions $\mathsf{Can}(\tD_f) : \List\ (\cN\times\cN)$, $\mathsf{CDom}(\tD_f), \mathsf{Dom}(\tD_f) : \List\ (\cN\times \cR)$ that compute the cancellation pairs of variables, codomain and domain.
    They are inductively defined by the structure of $\tD_f$:
    \begin{itemize}
        \item $\tD_f \equiv |i\>_r$: $\mathsf{Can}(\tD_f) = []$, $\mathsf{CDom}(\tD_f) = [(i,r)]$, $\mathsf{Dom}(\tD_f) = []$.
        \item $\tD_f \equiv {}_r\<i|$: $\mathsf{Can}(\tD_f) = []$, $\mathsf{CDom}(\tD_f) = []$, $\mathsf{Dom}(\tD_f) = [(i,r)]$.
        \item $\tD_f \equiv \tD_{1f}\otimes\tD_{2f}$: $\mathsf{Can}(\tD_f) = \mathsf{Can}(\tD_{1f}) ; \mathsf{Can}(\tD_{2f})$, $\mathsf{CDom}(\tD_f) = \mathsf{CDom}(\tD_{1f}) ; \mathsf{CDom}(\tD_{2f})$, $\mathsf{Dom}(\tD_f) = \mathsf{Dom}(\tD_{1f}) ; \mathsf{Dom}(\tD_{2f})$. (Here, $;$ is the concatenation of lists)
        \item $\tD_f \equiv \tD_{1f}\cdot\tD_{2f}$: recursively execute:
        \begin{enumerate}
            \item Set $L_{\rm can} = []$, $L_{\rm dom} = \mathsf{Dom}(\tD_{1f})$, $L_{\rm cdom} = \mathsf{CDom}(\tD_{2f})$.
            \item If there exists $(i_1,r)\in L_{\rm dom}$ and $(i_2,r)\in L_{\rm cdom}$, then $L_{\rm can} = L_{\rm can}; (i_1,i_2)$, $L_{\rm dom} = L_{\rm dom}\backslash(i_1,r)$ and $L_{\rm cdom} = L_{\rm cdom}\backslash(i_2,r)$. (Here, $\backslash$ means to remove the element from the list.)
            \item Repeat step (b) until it fails.
        \end{enumerate}
        Then, 
        $\mathsf{Can}(\tD_f) = \mathsf{Can}(\tD_{1f}) ; \mathsf{Can}(\tD_{2f}); L_{\rm can}$, 
        $\mathsf{CDom}(\tD_f) = \mathsf{CDom}(\tD_{1f}) ; L_{\rm cdom}$, $\mathsf{Dom}(\tD_f) = \mathsf{Dom}(\tD_{f}) ; L_{\rm dom}$. 
    \end{itemize}
    Suppose $\mathsf{CDom}(\tD_{f}) = \{(i_1,r_1),\cdots,(i_n,r_n)\}$ and $\mathsf{Dom}(\tD_{f}) = \{(i'_1,r'_1),\cdots,(i'_{n'},r'_{n'})\}$ where the list is sorted by a default order of variables in $\cR$. We define:
    $$D_f = \bigg(\prod_{(i,j)\in \mathsf{Can}(\tD_f)}\delta_{i,j}\bigg).\Big((|i_1\>\otimes\cdots\otimes|i_n\>) \cdot (\<i'_1|\otimes\cdots\otimes\<i'_{n'}|)\Big).$$
    If $\mathsf{Dom}$ or $\mathsf{CDom}$ is an empty list, the $D_f$ is just a Ket or Bra.
    Then we claim:
    \begin{itemize}
        \item If $E[\Gamma]\vdash \tD_f : \DType(s_1,s_2)$, then $\{r_1,\cdots,r_n\} = s_1$ and
        $\{r'_1,\cdots,r'_{n'}\} = s_2$.
        \item Set $R = ((\cdots(r_1,r_2),\cdots),r_n)$ and $R' = ((\cdots(r'_1,r'_2),\cdots),r'_{n'})$, then
        $$\tD_{f} = (D_f)_{R,R'}$$
    \end{itemize}
    \item For the obtained formula in (3), we replace each $\tD_f$ by $D_f$ which is computed by step (4). Note that $D_f$ is a Dirac term without labels, so the equivalence of two labelled Dirac terms is reduced to decide the equivalence of two Dirac terms. This is guaranteed if two labelled Dirac terms are of the same type (by typing rules).
\end{enumerate}

>>>>

@@

\begin{enumerate}
  \item For any register $R$ s.t. $E[\Gamma]\vdash R : \reg \sigma$, suppose $\var R = \{r_1,r_2,\cdots,r_n\}$, 
  we introduce variables $i_{r_k} : \Basis(\sigma_{r_k})$ with $r_k : \reg \sigma_{r_k} \in E$ for $k = 1,\cdots, n$. We reconstruct the basis $|i_R\>$ (which is of type $\KType(\sigma)$) and $\<i_R|$ (which is of type $\BType(\sigma)$) of $R$ by:
  \begin{itemize}
    \item $R = r_k$, $|i_R\> \triangleq |i_{r_k}\>$ and $\<i_R| \triangleq \<i_{r_k}|$;
    \item $R = (R_1,R_2)$: $|i_R\> \triangleq |i_{R_1}\> \otimes |i_{R_2}\>$ and $\<i_R| \triangleq \<i_{R_1}| \otimes \<i_{R_2}|$.
  \end{itemize}
    \item Expansion of all $K_R,B_R,O_{R,R'}$

    After the rewriting of this step, the expression we have is always in the following form:
    $$
    \Big(\sum_{i_1}\cdots\sum_{i_n} a.\tD_f\Big) + \cdots + \Big(\sum_{i_1}\cdots\sum_{i_{n'}} a.\tD_f\Big)
    $$
    where $\tD_f$ is generated by :
    $$\tD_f ::= |i\>_r \mid {}_r\<i| \mid \te\otimes\te \mid \te\cdot\te. $$
    \item
    Suppose $\mathsf{CDom}(\tD_{f}) = \{(i_1,r_1),\cdots,(i_n,r_n)\}$ and $\mathsf{Dom}(\tD_{f}) = \{(i'_1,r'_1),\cdots,(i'_{n'},r'_{n'})\}$ where the list is sorted by a default order of variables in $\cR$. We define:
    $$D_f = \bigg(\prod_{(i,j)\in \mathsf{Can}(\tD_f)}\delta_{i,j}\bigg).\Big((|i_1\>\otimes\cdots\otimes|i_n\>) \cdot (\<i'_1|\otimes\cdots\otimes\<i'_{n'}|)\Big).$$
    If $\mathsf{Dom}$ or $\mathsf{CDom}$ is an empty list, the $D_f$ is just a Ket or Bra.
    Then we claim:
    \begin{itemize}
        \item If $E[\Gamma]\vdash \tD_f : \DType(s_1,s_2)$, then $\{r_1,\cdots,r_n\} = s_1$ and
        $\{r'_1,\cdots,r'_{n'}\} = s_2$.
        \item Set $R = ((\cdots(r_1,r_2),\cdots),r_n)$ and $R' = ((\cdots(r'_1,r'_2),\cdots),r'_{n'})$, then
        $$\tD_{f} = (D_f)_{R,R'}$$
    \end{itemize}
    \item For the obtained formula in (3), we replace each $\tD_f$ by $D_f$ which is computed by step (4). Note that $D_f$ is a Dirac term without labels, so the equivalence of two labelled Dirac terms is reduced to decide the equivalence of two Dirac terms. This is guaranteed if two labelled Dirac terms are of the same type (by typing rules).
\end{enumerate}
